---
title: "Geographic processing"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  github_document:
    toc: true
always_allow_html: true
urlcolor: blue
---

*This is an adapted version of Growing Shade for use in Dane County, WI. While most of the comments/code remain the same, significant changes are included in italics. The original source code can be downloaded here: https://github.com/Metropolitan-Council/growing.shade*


```{r include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F,
                      cache = F)
library(dplyr); library(tidyr); library(readr); library(stringr); library(tibble)
library(tigris)
library(sf)
library(tidycensus)
library(here)

st_erase = function(x, y) st_difference(x, st_union(st_combine(y)))

```

# Process geographies

For the Twin Cities, Growing Shade nests block groups (the core level of analyses) into larger neighborhood and city-level geographies. This step is not easily applied to other regions, so will likely need to be specifically tailored if applying the methods elsewhere. 

**NOTE:** this script **DOES** rely on some parameters found inside the "global" `01_tutorial.Rmd` script, so please be sure to run that before running this script! It is okay if the tutorial script encounters an error and can't run all the way through, you'll still be saving information about which state/counties to use here!

### Neighborhoods and city levels

Since we're going to be making a map which shows census tracts, cities, or neighborhoods depending on the user input, a crosswalk needs to be made which relates block groups to the city and neighborhood levels.

If you can't download with a code-based method, download specific geographies, and put them in the data-raw folder. 

*For the Dane County region, neighborhoods are based on Madison neighborhood associations. A shapefile can be downloaded here:*
*https://data-cityofmadison.opendata.arcgis.com/datasets/cityofmadison::neighborhood-associations/explore?location=43.055113%2C-89.458982%2C13.52*
*Note that this file may have overlapping boundaries and need some clean-up*

Adjust the code below as necessary to ensure that both `nhood_geo` (neighborhoods) and `ctu_geo` (city/townships) have a column named `GEO_NAME` and `geometry`. For the neighborhood data, there should also be a `city` column (i.e., "Minneapolis" or "St. Paul" for the Twin Cities region). *Madison is the only area with neighborhood data for Dane County, but should still include the city column*

After the raw geographies are downloaded, then you need to make a crosswalk which relates block groups into neighborhoods and cities. For this step, it is useful to remove major river features (boundaries around rivers often are poorly aligned, removing rivers makes generating the crosswalk much cleaner). *I'm not certain whether this is an issue in Dane County, so I didn't remove rivers/lakes - something to look into further*. At least in the Twin Cities, several block groups legitimately do fall within multiple cities, so this step is a admittedly a bit complicated. A simpler alternative is to just use the city/township in which the majority of the block group falls.

If this section doesn't apply for other regions, it should be easy enough to remove elements in the user-interface of the application.


```{r nhood-geo}
# neighborhood
madison <- read_sf(paste0(here::here(), "/data-raw/madison_nhoods/nhoods.shp")) %>%
  mutate(Shape_Area = as.numeric(st_area(.))) 


nhood_geo <- bind_rows(madison) %>%
  select(GEO_NAME, city)

#shorten the neighborhood names by removing suffixes that are not necessary for our purposes

shorten_name <- function(x) { 
  find_i <- gregexpr('Neighborhood Association|Condominium Association|Condominum Association|Condominium|Homeowners Association|Community Association|Home Owners Association|Homeowners Assoc|Neighborhood As|Owners Assoc|Owners Association|Association', x)[[1]][1]
  if (find_i > 0) {
    return (substr(x, 1, find_i-2))
  } else {
    return(x)
  }
}

nhood_geo$GEO_NAME <- sapply(nhood_geo$GEO_NAME, shorten_name, USE.NAMES = FALSE)
```

While the download can be automated for the tiger file, it's not the most updated boundaries. This might work for updates for future years, but it doesn't have the new boundaries with incorporating the town of madison, so using  data from https://gis-countyofdane.opendata.arcgis.com/pages/boundaries for now


```{r ctu-geo}
ctu_geo <- read_sf(paste0(here::here(), "/data-raw/MunicipalBoundaries/Municipal_Boundaries.shp")) %>% 
  mutate(short = strsplit(NAME, " of ")) %>% 
  unnest_wider(short, names_sep = "_") %>% 
  group_by(short_2) %>% 
  mutate(n = n()) %>%
  #specify the type of municipality for any duplicate names
  mutate(NAME = case_when(n > 1 ~ paste0(short_2, " (", short_1,")"),
                          TRUE ~ short_2)) %>%
  ungroup() %>% 
  arrange(NAME) %>%
  rename(GEO_NAME = NAME) %>% 
  select(GEO_NAME, geometry) %>% 
  #convert back to a sf object
  st_as_sf() %>%
  #drop the z dimension - not sure why the data has it in the first place, but it causes errors later
  st_zm()

```

old code:

```{r ctu-geo old}
ctu_geo <- tigris::county_subdivisions(
  state = "WI",
  county = c("Dane"),
  class = "sf"
) %>%
  mutate(NAME = case_when(LSAD == 44 ~  paste(NAME, "Twp."), 
                          LSAD == 46 ~ paste(NAME, "(unorg.)"),
                          TRUE ~ NAME)) %>%
  group_by(NAME) %>%
  mutate(n = n()) %>%
  left_join(st_drop_geometry(county_outline) %>%
              transmute(COUNTYFP = COUNTYFP, 
                        CONAME = NAME)) %>%
  mutate(NAME = case_when(n > 1 & LSAD != 25 ~ paste0(NAME, " - ", CONAME, " Co."), #cities dont get merged 
                          TRUE ~ NAME)) %>%
  group_by(NAME) %>%
  summarise() %>%
  # summarize(geometry = st_union(geom)) %>%
  arrange(NAME) %>%
  rename(GEO_NAME = NAME)
```

```{r ctu-nhood-crosswalk}

# fxns to make easy -----
# find crosswalks
find_crosswalks <- function(x) {
  crosswalk <- x %>%
  st_transform(26915) %>%
  #st_buffer(-150) %>% #buffer the perimeter of the geography - this line is causing issues with neighborhoods that are smaller than the buffer, not sure why the buffer is included in the first place
  st_intersection(bg_geo %>% 
                    dplyr::select(GEOID) %>%
                    rename(bg_id = GEOID) %>%
                    st_transform(26915)) %>%
  st_drop_geometry() %>% 
  #filter out the two block groups that make up lake mendota and monona
  filter(bg_id != "550259917030" & bg_id != "550259917020")
    
  return(crosswalk)
}

ctu_crosswalk <- find_crosswalks(ctu_geo)

nhood_crosswalk <- find_crosswalks(nhood_geo)

wide_ctu_crosswalk <- ctu_crosswalk %>%
    aggregate(GEO_NAME ~ bg_id, paste, collapse = ", ") %>%
  rename(jurisdiction = GEO_NAME)

  save(ctu_crosswalk, nhood_crosswalk, wide_ctu_crosswalk, file = paste0(here::here(), "/data-raw/geography_data.rda"))

```
