---
title: "Tree canopy processing"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  github_document:
    toc: true
always_allow_html: true
urlcolor: blue
---

*This is an adapted version of Growing Shade for use in Dane County, WI. While most of the comments/code remain the same, significant changes are included in italics. The original source code can be downloaded here: https://github.com/Metropolitan-Council/growing.shade*

For the Twin Cities, Growing Shade nests block groups (the core level of analyses) into larger neighborhood and city-level geographies. This step is not easily applied to other regions, so will likely need to be specifically tailored if applying the methods elsewhere. 

**NOTE:** this script **DOES** rely on some parameters found inside the "global" `01_tutorial.Rmd` script, so please be sure to run that before running this script! It is okay if the tutorial script encounters an error and can't run all the way through, you'll still be saving information about which state/counties to use here!

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F,
                      cache = F)
library(dplyr); library(tidyr); library(readr); library(stringr); library(tibble); library(ggplot2)
library(tigris)
library(sf)
library(tidycensus)
library(ggbeeswarm)
library(RSocrata)
library(here)
library(terra)

st_erase = function(x, y) st_difference(x, st_union(st_combine(y)))
`%not_in%` <- Negate(`%in%`)
```



# Remote sensing data 

These items come out of Google Earth Engine analyses. Some files need to be created/exported (and then imported into GEE), and all GEE exports need to be pulled in to this code.


## Tree canopy data

There are many ways of measuring the tree canopy, and each method has pros/cons. 


### Calibrate tree canopy coverage

Growing Shade prioritizes temporal accuracy for tree canopy data. While some trade-offs come with prioritizing temporal accuracy over spatial accuracy, it is essential for this project to capture on-the-ground, real-time dynamics of how Emerald Ash Borer, development patterns, and recent tree planting programs among others are changing the tree canopy.

Sentinel-2 is currently the most spatially accurate and publicly accessible remote sensing platform. However, the 10 meter squared spatial resolution of Sentinel is larger than a lot of individual tree canopies. In exploring the data, it appears as if the canopy coverage from Sentinel is little higher than what it should be (based on aerial imagery). I've chosen to calibrate the Sentinel data with the (outdated) UMN 1 meter squared land use file.

I created a grid (n = 1015) across the region, and created a model to compare the amount of trees detected with the Sentinel 2 and with the UMN 1 meter data set (for the later, the tree area is the summation of all areas which identified as coniferous trees, deciduous trees, and forested wetland).

The final calibration coefficient (at least for 2021 tree canopy) is 0.885246, meaning that Sentinel sees about 11% more trees in areas. Another way to think about this is that Sentinel detects area with at least 89% tree canopy coverage (i.e., if Sentinel sees 1,000 acres of trees, UMN sees more like 885 acres).

In other temperate, upper Midwest areas, the .88 coefficient is probably sufficient. Otherwise, figuring out and adjusting a calibration coefficient for another area likely requires some pretty bespoke analyses and has specific data set needs which may not be widely available. 

*I haven't done the same analysis for Dane County yet, so used the same 0.885246 coefficient*


```{r, calibrate-tree-cover}

# ######
# # Create a gridded area across the region to calibrate sentinel tree cover with 1m2 land cover tree data
# # In most instances, there is no need to run this gridding step more than once
# #####
# wholearea <- metc_region %>%
#   summarise(st_union(.))
# 
# # make a equal area grid; there are 704 tracts, so I want to make at least 1000 grids I think?
# g = st_make_grid(wholearea,
#                  n = c(36, 36)) %>% 
#   st_intersection(wholearea) 
# 
# geometry = st_sfc(lapply(1:length(g), function(x) st_geometrycollection()))
# df <- st_sf(id = 1:length(g), geometry = g)
# 
# # ggplot() +
# #   geom_sf(data = wholearea) +
# #   geom_sf(data = df,
# #           fill = "transparent")
# 
# sf::st_write(df, "~/Documents/GitHub/planting.shade/storymap-info/shapefiles/metc_grid.shp", append = FALSE)

calibrate_trees <- read_csv(paste0(here::here(), "/data-raw/UMNTreeAcres_metcgrid_scale1_year2021.csv")) %>% 
  rename(umn = `1`) %>%
  full_join(read_csv(paste0(here::here(),"/data-raw/TreeAcres_metcgrid_year2021.csv")) %>%
              rename(sentinel = `1`),
            by = 'id')

calibrate_lm <- (lm(umn ~ sentinel, data = calibrate_trees))
calibrate_lm2 <- (lm(umn ~ 0 + sentinel, data = calibrate_trees))
calibrate_lm3 <- (lm(umn ~ I(sentinel ^ 2), data = calibrate_trees))
calibrate_lm4 <- (lm(log(umn) ~ sentinel, data = calibrate_trees))
anova(calibrate_lm, calibrate_lm2, calibrate_lm3, calibrate_lm4) # the middle model is best!

# AIC(calibrate_lm); AIC(calibrate_lm2); AIC(calibrate_lm3); AIC(calibrate_lm4)

summary(calibrate_lm2)$r.squared # r2
summary(calibrate_lm2)$coefficients[,4] # p-value

calib_coeff <- summary(calibrate_lm2)$coefficients[,1] # coefficient

  # save(file = paste0(here::here(), "/data-raw/calib_coeff.rda"), calib_coeff)


calibrate_trees %>%
  ggplot(aes(x = (umn), y = (sentinel * calib_coeff))) +
  geom_point(alpha = .5) +
  geom_abline(slope=1, intercept=0, col = 'blue', lwd = 1) +
  theme_minimal()  +
  labs( x = "UMN tree acres", y = "Calibrated Sentinel tree acres")
```

### Process tree canopy cover at various geographies

-   process GEE code to link canopy with geography
    -   GEE data is in repo "users/ehe/MetCoucil/GrowingShade_CanopyCoverage"
    -   <https://code.earthengine.google.com/a0da66053ecb26b668df4297c4ebed59>

*The GEE output from the link above didn't seem to output a file in the correct format to be processed by the original code. I made changes to the code below and the GEE script: https://code.earthengine.google.com/d34583e1d03f11b2c02ec7a3a0e2592d*




```{r canopy-processing bg}
#block groups

#just using the same coefficient for the twin cities for now, will potentially calculate for Madison later
calib_coeff <- .88

bg_canopy <- read_csv(paste0(here::here(), "/data-raw/DC_TreeAcres_blockgroups_year2021.csv")) %>% 
  arrange(GEOID)

bg_geo <- arrange(bg_geo, GEOID)
bg_canopy$ALAND <- bg_geo$ALAND

bg <- vect(bg_geo)
#calculate total area of all block groups in meters
total_area <- expanse(bg,"m")

bg_canopy <- bg_canopy %>% 
  rename('trees' = '1', 'treeless' = '0') %>% 
  #convert square acres (GEE output) to meters
  mutate(trees = trees*4046.856422,
         treeless = total_area - trees,
         treeless = if_else(ALAND == 0, 0, as.numeric(treeless)),
         trees = if_else(ALAND == 0, 0, as.numeric(trees)),
         canopy_percent = trees / (trees + treeless) * calib_coeff, 
         canopy_percent = if_else(is.na(canopy_percent), 0, canopy_percent)) %>%
    filter(ALAND != 0) %>% #new 2020 block groups sometimes are only water
    mutate(avgcanopy = mean(canopy_percent)) %>%
    select(-ALAND, -trees, -treeless) %>% 
    rename(bg_id = GEOID) %>%
    transform(bg_id = as.character(bg_id)) %>% 
    #filter out the two block groups that make up lake mendota and monona
    filter(bg_id != "550259917030" & bg_id != "550259917020")

```

```{r canopy-processing ctu}
#cities

ctu_list_raw <- read_csv(paste0(here::here(), "/data-raw/DC_TreeAcres_ctu_year2021.csv")) %>% 
  arrange(GEO_NAME)

ctu_geo <- st_read(paste0(here::here(), "/data-raw/gee_canopy_data/cities/ctu.shp")) %>% 
  arrange(GEO_NAME)

ctu <- vect(ctu_geo)
#calculate total area of all cities in meters
total_area <- expanse(ctu,"m")

  

ctu_list_raw <- ctu_list_raw %>% 
  rename('trees' = '1', 'treeless' = '0') %>% 
  #convert square acres (GEE output) to meters
  mutate(trees = trees*4046.856422,
         treeless = total_area - trees,
         canopy_percent = trees / (trees + treeless) * calib_coeff, 
         canopy_percent = if_else(is.na(canopy_percent), 0, canopy_percent)) %>%
    mutate(avgcanopy = mean(canopy_percent)) %>%
    select(-trees, -treeless) %>%
  full_join(left_join(ctu_crosswalk, bg_canopy) %>% 
  group_by(GEO_NAME) %>%
  summarise(
    min = round(min(canopy_percent) * 100, 1),
    max = round(max(canopy_percent) * 100, 1),
    n_blockgroups = n()
  )) %>%
  arrange(GEO_NAME) %>%
  full_join(ctu_geo) %>%
  st_as_sf()
```

```{r canopy-processing nhood}
#neighborhoods
nhood_list <- read_csv(paste0(here::here(), "/data-raw/DC_TreeAcres_nhood_year2021.csv")) %>% 
  arrange(GEO_NAME)

nhood_geo <- st_read(paste0(here::here(), "/data-raw/gee_canopy_data/nhood/nhood.shp")) %>% 
  arrange(GEO_NAME)

nhood <- vect(nhood_geo)
#calculate total area of all cities in meters
total_area <- expanse(nhood,"m")

nhood_list_raw <- nhood_list %>% 
  rename('trees' = '1', 'treeless' = '0') %>% 
  #convert square acres (GEE output) to meters
  mutate(trees = trees*4046.856422,
         treeless = total_area - trees,
         canopy_percent = trees / (trees + treeless) * calib_coeff, 
         canopy_percent = if_else(is.na(canopy_percent), 0, canopy_percent)) %>%
    mutate(avgcanopy = mean(canopy_percent)) %>%
    select(-trees, -treeless) %>% 
  full_join(left_join(nhood_crosswalk, bg_canopy) %>% 
  group_by(GEO_NAME, city) %>%
  summarise(
    min = round(min(canopy_percent) * 100, 1),
    max = round(max(canopy_percent) * 100, 1),
    n_blockgroups = n()
  )) %>%
  full_join(nhood_geo) %>%
  st_as_sf()
```



## Greenness (NDVI) Data

We do this for all land (no water!) and non-cultivated land (excluding crops/ag land).

```{r ndvi_bgs}
ndvi_uncultivated <- 
  read_csv(paste0(here::here(), "/data-raw/DC_uncultivated_meanNDVI_bg_year2021.csv"),
                    na = "No data",
                       col_types = cols(GEOID = "c", `system:index` = "c", Year = 'd',  `.geo` = 'c')) %>%
  rename(GEOID = GEOID, ndvi_uncultivated = ndvi)

ndvi_allland <- 
  read_csv(paste0(here::here(), "/data-raw/DC_land_meanNDVI_bg_year2021.csv"),
                    na = "No data",
                       col_types = cols(GEOID = "c", `system:index` = "c", Year = 'd',  `.geo` = 'c')) %>%
  rename(GEOID = GEOID, ndvi_land = ndvi)
  
bg_ndvi <- ndvi_uncultivated %>%
  dplyr::select(GEOID, ndvi_uncultivated) %>%
  full_join(ndvi_allland %>%
  dplyr::select(GEOID, ndvi_land)) %>%
    rename(bg_id = GEOID)

```

## Save data


```{r}
  save(bg_canopy, bg_ndvi, ctu_list_raw, nhood_list_raw, file = paste0(here::here(), "/data-raw/canopy_data.rda"))

```


## Climate vulnerability processing

*The original code used data specific to the Twin Cities region. This code was written to create an equivalent  climate vulnerability data set for Dane County.*

*This code processes two data sets that should be in data-raw:*
*1. FEMA primary flood plains: https://geodata.wisc.edu/catalog/E2CE7AA7-7E6B-4E6C-9237-DA55D4AB69CC (TODO - this data set had extra information, can the floodplains be downloaded directly from FEMA?)*
*2. Average land surface temperature during a heatwave, derived from MODIS data and google earth engine:
https://code.earthengine.google.com/010f56d1c02224980092e2f6711cb186*


```{r environmental-processing}
# Block group geometries are downloaded in 01_tutorial.Rmd
#TODO - am I including lake areas in the temperature?
bg <- vect(bg_geo)

#the average land surface temperature
avgTemp <- rast(paste0(here::here(), "/data-raw/heatwave_mean_lst.tif"))
#convert temperatures from Kelvin to Fahrenheit
avgTemp$tempF <- (avgTemp$tempScale - 273.15) * 9/5 + 32
#calculate the mean temperature within each block group
bg_temp <- zonal(avgTemp$tempF, bg, "mean", na.rm=TRUE)

#primary flood plain
flood <- vect(paste0(here::here(), "/data-raw/fema_floodplain/2023_DC_FEMA_floodplain.shp"))
#reproject
flood <- project(flood, bg)

clim_vul <- data.frame(matrix(nrow = 353, ncol = 3))
columns= c("BG55", "AVG_TEMP", "PRIM_FLOOD")
colnames(clim_vul) = columns

clim_vul$BG55 <- bg$GEOID
clim_vul$AVG_TEMP <- bg_temp$tempF

#calculate the percentage of land susceptible to flooding withing each block group
for (i in 1:nrow(bg)) {
  clip <- crop(flood, bg[i,])
  perc <- expanse(clip)/expanse(bg[i,])
  if (length(perc > 0)) {
    clim_vul[i,3] <- perc
  } else {
    clim_vul[i,3] <- 0
  }
}

#write csv
write.csv(clim_vul, paste0(here::here(), "/data-raw/CLIMATE_BG55.csv"), row.names = FALSE)
```


##Calculate % of total canopy for land use categories

TODO - clean up this code - should definitely extract a lot of it into functions
```{r}
#load datasets
tree_mask <- rast(paste0(here::here(), "/data-raw/tree_mask_2021.tif"))
lui <- rast(paste0(here::here(), "/data-raw/LUI_2020_Dane_County/lui_2020_rasterized_10m.tif"))

#create block group canopy lu dataframe
bg_canopy_lu <- data.frame(matrix(nrow = 0, ncol=17))
colnames(bg_canopy_lu) = c("GEO_NAME", unlist(unique(lui$LANDUSE_CA)))
bg_canopy_lu$GEO_NAME = as.character(bg_canopy_lu$GEO_NAME)

#calculate the % of total canopy on each land use category for block groups
for (i in 1:nrow(bg_geo)) {
  print(paste(i, "/", nrow(bg_geo)))
  #grab the vector data for the area
  aoi <- bg_geo[i,]
  #mask the tree and lui rasters to the correct area
  tree_mask_crop <- crop(tree_mask, aoi, mask=TRUE)
  lui_crop <- crop(lui, aoi, mask=TRUE)
  
  #find the amount of canopy on each type of land
  cur_canopy <- zonal(tree_mask_crop, lui_crop, fun="sum", na.rm=TRUE)
  #Subtract any canopy on water (due to sentinel 10m resolution) from the total %s calculated
  if(length(filter(cur_canopy, LANDUSE_CA == 'Water')$trees) != 0) {
    total <- sum(cur_canopy$trees) - filter(cur_canopy, LANDUSE_CA == 'Water')$trees
  } else {
    total <- sum(cur_canopy$trees)
  }
  
  cur_canopy <- cur_canopy %>%
    mutate(perc = trees/total*100) %>%
    select(-trees) %>%
    spread(LANDUSE_CA, perc) %>%
    mutate(GEO_NAME = aoi$GEOID)
  #add to the big dataframe
  bg_canopy_lu <- bind_rows(bg_canopy_lu, cur_canopy)
}

#post-processing
#filter out lake mendota and monona block groups
bg_canopy_lu <- filter(bg_canopy_lu, GEO_NAME != "550259917030") %>% 
  filter(GEO_NAME != "550259917020") %>% 
  #remove misclassified water data
  select(-Water) %>% 
  #set anything with less than 5% coverage as NA
  mutate(across(Agriculture:Woodlands, round)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
#delete any columns with only NA values
bg_canopy_lu <- bg_canopy_lu[, colSums(is.na(bg_canopy_lu)) != nrow(bg_canopy_lu)]


#calculate the % of total canopy on each land use category for neighborhoods

#create neighborhood group canopy lu dataframe
nhood_canopy_lu <- data.frame(matrix(nrow = 0, ncol=17))
colnames(nhood_canopy_lu) = c("GEO_NAME", unlist(unique(lui$LANDUSE_CA)))
nhood_canopy_lu$GEO_NAME = as.character(nhood_canopy_lu$GEO_NAME)

nhood_geo_proj <- st_transform(nhood_geo, 4269)

for (i in 1:nrow(nhood_geo_proj)) {
  print(paste(i, "/", nrow(nhood_geo_proj)))
  #grab the vector data for the area
  aoi <- nhood_geo_proj[i,]
  #mask the tree and lui rasters to the correct area
  tree_mask_crop <- crop(tree_mask, aoi, mask=TRUE)
  lui_crop <- crop(lui, aoi, mask=TRUE)
  
  #find the amount of canopy on each type of land
  cur_canopy <- zonal(tree_mask_crop, lui_crop, fun="sum", na.rm=TRUE)
  #Subtract any canopy on water (due to sentinel 10m resolution) from the total %s calculated
  if(length(filter(cur_canopy, LANDUSE_CA == 'Water')$trees) != 0) {
    total <- sum(cur_canopy$trees) - filter(cur_canopy, LANDUSE_CA == 'Water')$trees
  } else {
    total <- sum(cur_canopy$trees)
  }
  
  cur_canopy <- cur_canopy %>%
    mutate(perc = trees/total*100) %>%
    select(-trees) %>%
    spread(LANDUSE_CA, perc) %>%
    mutate(GEO_NAME = aoi$GEO_NAME)
  #add to the big dataframe
  nhood_canopy_lu <- bind_rows(nhood_canopy_lu, cur_canopy)
}

#post-processing
nhood_canopy_lu <- nhood_canopy_lu %>% 
  select(-Water) %>% 
  mutate(across(Agriculture:Woodlands, round)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
#delete any columns with only NA values
nhood_canopy_lu <- nhood_canopy_lu[, colSums(is.na(nhood_canopy_lu)) != nrow(nhood_canopy_lu)]

  
#calculate the % of total canopy on each land use category for ctu

#create ctu group canopy lu dataframe
ctu_canopy_lu <- data.frame(matrix(nrow = 0, ncol=17))
colnames(ctu_canopy_lu) = c("GEO_NAME", unlist(unique(lui$LANDUSE_CA)))
ctu_canopy_lu$GEO_NAME = as.character(ctu_canopy_lu$GEO_NAME)

ctu_geo_proj <- st_transform(ctu_geo, 4269)

for (i in 1:nrow(ctu_geo_proj)) {
  print(paste(i, "/", nrow(ctu_geo_proj)))
  #grab the vector data for the area
  aoi <- ctu_geo_proj[i,]
  #mask the tree and lui rasters to the correct area
  tree_mask_crop <- crop(tree_mask, aoi, mask=TRUE)
  lui_crop <- crop(lui, aoi, mask=TRUE)
  
  #find the amount of canopy on each type of land
  cur_canopy <- zonal(tree_mask_crop, lui_crop, fun="sum", na.rm=TRUE)
  #Subtract any canopy on water (due to sentinel 10m resolution) from the total %s calculated
  if(length(filter(cur_canopy, LANDUSE_CA == 'Water')$trees) != 0) {
    total <- sum(cur_canopy$trees) - filter(cur_canopy, LANDUSE_CA == 'Water')$trees
  } else {
    total <- sum(cur_canopy$trees)
  }
  
  cur_canopy <- cur_canopy %>%
    mutate(perc = trees/total*100) %>%
    select(-trees) %>%
    spread(LANDUSE_CA, perc) %>%
    mutate(GEO_NAME = aoi$GEO_NAME)
  #add to the big dataframe
  ctu_canopy_lu <- bind_rows(ctu_canopy_lu, cur_canopy)
}

#post-processing
ctu_canopy_lu <- ctu_canopy_lu %>% 
  select(-Water) %>%
  mutate(across(Agriculture:Woodlands, round)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
  #delete any columns with only NA values
ctu_canopy_lu <- ctu_canopy_lu[, colSums(is.na(ctu_canopy_lu)) != nrow(ctu_canopy_lu)]
  
#save all 3 files
save(bg_canopy_lu, ctu_canopy_lu, nhood_canopy_lu, file = paste0(here::here(), "/data-raw/landuse_data.rda"))

```


##Create a map of planting opportunities

```{r}
ndvi <- rast(paste0(here::here(), "/data-raw/summer_ndvi_2021.tif"))
tree_mask <- rast(paste0(here::here(), "/data-raw/tree_mask_2021.tif"))

#arbitrary ndvi threshold - the goal is to filter out anything that's not vegetation
#I decided on .7 by just looking at some grassy open areas like parks, but it probably deserves a more systematic method
ndvi_thresh <- .7
opp_map <- crop(ndvi, tree_mask, mask=TRUE) %>% 
  classify(matrix(c(0, ndvi_thresh, NA), ncol=3, byrow=TRUE), include.lowest=TRUE) %>% 
  mask(tree_mask, maskvalues=1) %>%
  #replace ndvi values with a boolean value
  classify(matrix(c(0, 1, 1), ncol=3, byrow=TRUE), include.lowest=TRUE)

writeRaster(opp_map, paste0(here::here(), "/data-raw/opp_map_.7ndvi.tif"), filetype = "GTiff", datatype = "INT1U")
```

##Calculate % of opportunities for land use categories


```{r}
#load datasets
opp_map <- rast(paste0(here::here(), "/data-raw/opp_map_.7ndvi.tif"))
lui <- rast(paste0(here::here(), "/data-raw/LUI_2020_Dane_County/lui_2020_rasterized_10m.tif"))

#aoi <- ctu_geo_proj[10,]
find_lui <- function(aoi, geo) {
  #mask the opp and lui rasters to the correct area
  opp_crop <- crop(opp_map, aoi, mask=TRUE)
  lui_crop <- crop(lui, aoi, mask=TRUE)
  
  #find the amount of canopy on each type of land
  opp_lui <- zonal(opp_crop, lui_crop, fun="sum", na.rm=TRUE) %>% 
    rename(pixels = `uncultivated summer max NDVI`) %>% 
    drop_na()
  #Subtract any opportunities on water (due to sentinel 10m resolution) from the total %s calculated
  if(length(filter(opp_lui, LANDUSE_CA == 'Water')$pixels) != 0) {
    total <- sum(opp_lui$pixels) - filter(opp_lui, LANDUSE_CA == 'Water')$pixels
  } else {
    total <- sum(opp_lui$pixels)
  }
  
  opp_lui <- opp_lui %>%
    mutate(perc = pixels/total*100) %>% 
    #mutate('Total Opportunity' = units::drop_units(total*100/st_area(aoi))*100) %>% 
    mutate(Acres = sum(pixels)*100/4046.856422) %>% 
    select(-pixels) %>% 
    spread(LANDUSE_CA, perc)
  
  if (geo == "bg") {  
    opp_lui <- mutate(opp_lui, GEO_NAME = aoi$GEOID)
  } else if (geo == "ctu" | geo == "nhood") {
    opp_lui <- mutate(opp_lui, GEO_NAME = aoi$GEO_NAME)
  } 
  
  return (opp_lui)
}



#calculate the % of total canopy on each land use category for block groups

#create block group canopy lu dataframe
bg_opp_lu <- data.frame(matrix(nrow = 0, ncol=18))
colnames(bg_opp_lu) = c("GEO_NAME", "Acres", unlist(unique(lui$LANDUSE_CA)))
bg_opp_lu$GEO_NAME = as.character(bg_opp_lu$GEO_NAME)
#bg_geo_proj <- st_transform(bg_geo, 4269)

for (i in 1:nrow(bg_geo)) {
  print(paste(i, "/", nrow(bg_geo)))
  #add to the big dataframe
  bg_opp_lu <- bind_rows(bg_opp_lu,  find_lui(bg_geo[i,], "bg"))
}

#post-processing
#filter out lake mendota and monona block groups
bg_opp_lu <- filter(bg_opp_lu, GEO_NAME != "550259917030") %>% 
  filter(GEO_NAME != "550259917020") %>% 
  #remove misclassified water data
  select(-Water) %>% 
  #set anything with less than 5% coverage as NA
  mutate(across(Agriculture:Woodlands, round)) %>%
  mutate(Acres = round(Acres, 2)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
#delete any columns with only NA values
bg_opp_lu <- bg_opp_lu[, colSums(is.na(bg_opp_lu)) != nrow(bg_opp_lu)]


#calculate the % of total canopy on each land use category for neighborhoods

#create neighborhood canopy lu dataframe
nhood_opp_lu <- data.frame(matrix(nrow = 0, ncol=18))
colnames(nhood_opp_lu) = c("GEO_NAME", "Acres", unlist(unique(lui$LANDUSE_CA)))
nhood_opp_lu$GEO_NAME = as.character(nhood_opp_lu$GEO_NAME)
nhood_geo_proj <- st_transform(nhood_geo, 4269)

for (i in 1:nrow(nhood_geo_proj)) {
  print(paste(i, "/", nrow(nhood_geo_proj)))
  #add to the big dataframe
  nhood_opp_lu <- bind_rows(nhood_opp_lu,  find_lui(nhood_geo_proj[i,], "nhood"))
}

#post-processing
nhood_opp_lu <- nhood_opp_lu %>% 
  #remove misclassified water data
  select(-Water) %>% 
  #set anything with less than 5% coverage as NA
  mutate(across(Agriculture:Woodlands, round)) %>%
  mutate(Acres = round(Acres, 2)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
#delete any columns with only NA values
nhood_opp_lu <- nhood_opp_lu[, colSums(is.na(nhood_opp_lu)) != nrow(nhood_opp_lu)]


#calculate the % of total canopy on each land use category for neighborhoods

#create neighborhood canopy lu dataframe
ctu_opp_lu <- data.frame(matrix(nrow = 0, ncol=18))
colnames(ctu_opp_lu) = c("GEO_NAME", "Acres", unlist(unique(lui$LANDUSE_CA)))
ctu_opp_lu$GEO_NAME = as.character(ctu_opp_lu$GEO_NAME)
#ctu_geo_proj <- st_transform(ctu_geo, 4269)

for (i in 1:nrow(ctu_geo_proj)) {
  print(paste(i, "/", nrow(ctu_geo_proj)))
  #add to the big dataframe
  ctu_opp_lu <- bind_rows(ctu_opp_lu,  find_lui(ctu_geo_proj[i,], "ctu"))
}

#post-processing
ctu_opp_lu <- ctu_opp_lu %>% 
  #remove misclassified water data
  select(-Water) %>% 
  #set anything with less than 5% coverage as NA
  mutate(across(Agriculture:Woodlands, round)) %>%
  mutate(Acres = round(Acres, 2)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
#delete any columns with only NA values
ctu_opp_lu <- ctu_opp_lu[, colSums(is.na(ctu_opp_lu)) != nrow(ctu_opp_lu)]

#save all 3 files
save(bg_opp_lu, ctu_opp_lu, nhood_opp_lu, file = paste0(here::here(), "/data-raw/landuse_opp_data.rda"))

```

Experimental 'zoom-in' planner:

```{r}

#load datasets
ndvi <- rast(paste0(here::here(), "/data-raw/summer_ndvi_2021.tif"))
tree_mask <- rast(paste0(here::here(), "/data-raw/tree_mask_2021.tif"))
#lui2 <- vect(paste0(here::here(), "/data-raw/LUI_2020_Dane_County/LUI_2020_Dane_County_Project.shp"))
lui <- rast(paste0(here::here(), "/data-raw/LUI_2020_Dane_County/lui_2020_rasterized_10m.tif"))

#create block group canopy lu dataframe
bg_canopy_lu <- data.frame(matrix(nrow = 0, ncol=17))
colnames(bg_canopy_lu) = c("GEO_NAME", unlist(unique(lui$LANDUSE_CA)))
bg_canopy_lu$GEO_NAME = as.character(bg_canopy_lu$GEO_NAME)

#bg_canopy_lu$GEO_NAME = bg_geo$GEOID
#lui2 <- st_read(paste0(here::here(), "/data-raw/LUI_2020_Dane_County/LUI_2020_EXPORT_for_CARPC.gdb"))
aoi <- filter(bg_geo, GEOID == "550250132011")
#aoi <- sf::st_transform(filter(nhood_geo, GEO_NAME == "Vilas"), 4269)
#aoi <- sf::st_transform(filter(ctu_geo, GEO_NAME == "Belleville"), 4269)
#aoi <- vect(aoi)

ndvi_crop <- crop(ndvi, aoi)
ndvi_crop <- crop(ndvi, aoi, mask=TRUE)

tree_mask_crop <- crop(tree_mask, aoi)
lui_crop <- crop(lui, ndvi_crop, ext=FALSE)
r <- rasterize(lui_crop, ndvi_crop, "LANDUSE_CA") %>% 
  #mask out areas outside of the boundary
  mask(aoi)
#tst <- sf::st_transform(filter(nhood_geo, GEO_NAME == "Vilas"), 4269)
#tst <- sf::st_transform(filter(ctu_geo, GEO_NAME == "Albion"), 4269)

#ndvi_crop <- rast(st_crop(ndvi, aoi))
#tree_mask_crop <- rast(st_crop(tree_mask, aoi))
#use a 3x3 moving window to find larger areas of deficient canopy
#this means each pixel measures about how much canopy is present in .5 mile 
tree_mask_focal <- focal(tree_mask_crop, 3, sum)
#arbitrary ndvi threshold - the goal is to filter out anything that's not vegetation
#I decided on .7 by just looking at some grassy open areas like parks, but it probably deserves a more systematic method
ndvi_thresh <- .7
ndvi_crop <- classify(ndvi_crop, matrix(c(0, ndvi_thresh, 0), ncol=3, byrow=TRUE), include.lowest=TRUE)
#mask out areas that already have trees
opp_map <- mask(tree_mask_focal, tree_mask_crop, maskvalues=1) %>% 
  #mask out areas outside of the boundary
  mask(aoi) %>% 
  #mask out areas that aren't greenspace
  mask(ndvi_crop, maskvalues=0) %>% 
  mask(ndvi_crop, maskvalues=NA) %>% 
  #make a simplified summary
  classify(matrix(c(0, 2, 1, 2, 5, 2, 5, 8, 3), ncol=3, byrow=TRUE), include.lowest=TRUE)

total <- units::drop_units(st_area(aoi))
  
  # technically more exact since it uses the shapefile boundaries instead of the raster, but practically made less than 1% difference and was much slower
  # cur_canopy <- extract(tree_mask_crop, lui_crop, fun="sum", na.rm=TRUE, exact=TRUE)
  # cur_canopy$class <- lui_crop$LANDUSE_CA
  # cur_canopy <- cur_canopy %>%
  #   drop_na() %>%
  #   group_by(class) %>%
  #   summarise(total = sum(trees)) %>%
  #   mutate(perc = total/sum(total)*100) %>%
  #   select(-total) %>%
  #   spread(class, perc) %>%
  #   mutate(GEO_NAME = aoi$GEOID)
  

#calculate the % of total canopy on each land use category for block groups
for (i in 1:nrow(bg_geo)) {
  print(paste(i, "/", nrow(bg_geo)))
  #grab the vector data for the area
  aoi <- bg_geo[i,]
  #mask the tree and lui rasters to the correct area
  tree_mask_crop <- crop(tree_mask, aoi, mask=TRUE)
  lui_crop <- crop(lui, aoi, mask=TRUE)
  
  #find the amount of canopy on each type of land
  cur_canopy <- zonal(tree_mask_crop, lui_crop, fun="sum", na.rm=TRUE)
  #Subtract any canopy on water (due to sentinel 10m resolution) from the total %s calculated
  if(length(filter(cur_canopy, LANDUSE_CA == 'Water')$trees) != 0) {
    total <- sum(cur_canopy$trees) - filter(cur_canopy, LANDUSE_CA == 'Water')$trees
  } else {
    total <- sum(cur_canopy$trees)
  }
  
  cur_canopy <- cur_canopy %>%
    mutate(perc = trees/total*100) %>%
    select(-trees) %>%
    spread(LANDUSE_CA, perc) %>%
    mutate(GEO_NAME = aoi$GEOID)
  #add to the big dataframe
  bg_canopy_lu <- bind_rows(bg_canopy_lu, cur_canopy)
}

#post-processing
#filter out lake mendota and monona block groups
bg_canopy_lu <- filter(bg_canopy_lu, GEO_NAME != "550259917030") %>% 
  filter(GEO_NAME != "550259917020") %>% 
  #remove misclassified water data
  select(-Water) %>% 
  #set anything with less than 5% coverage as NA
  mutate(across(Agriculture:Woodlands, round)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
#delete any columns with only NA values
bg_canopy_lu <- bg_canopy_lu[, colSums(is.na(bg_canopy_lu)) != nrow(bg_canopy_lu)]


#calculate the % of total canopy on each land use category for neighborhoods

#create neighborhood group canopy lu dataframe
nhood_canopy_lu <- data.frame(matrix(nrow = 0, ncol=17))
colnames(nhood_canopy_lu) = c("GEO_NAME", unlist(unique(lui$LANDUSE_CA)))
nhood_canopy_lu$GEO_NAME = as.character(nhood_canopy_lu$GEO_NAME)

nhood_geo_proj <- st_transform(nhood_geo, 4269)

for (i in 1:nrow(nhood_geo_proj)) {
  print(paste(i, "/", nrow(nhood_geo_proj)))
  #grab the vector data for the area
  aoi <- nhood_geo_proj[i,]
  #mask the tree and lui rasters to the correct area
  tree_mask_crop <- crop(tree_mask, aoi, mask=TRUE)
  lui_crop <- crop(lui, aoi, mask=TRUE)
  
  #find the amount of canopy on each type of land
  cur_canopy <- zonal(tree_mask_crop, lui_crop, fun="sum", na.rm=TRUE)
  #Subtract any canopy on water (due to sentinel 10m resolution) from the total %s calculated
  if(length(filter(cur_canopy, LANDUSE_CA == 'Water')$trees) != 0) {
    total <- sum(cur_canopy$trees) - filter(cur_canopy, LANDUSE_CA == 'Water')$trees
  } else {
    total <- sum(cur_canopy$trees)
  }
  
  cur_canopy <- cur_canopy %>%
    mutate(perc = trees/total*100) %>%
    select(-trees) %>%
    spread(LANDUSE_CA, perc) %>%
    mutate(GEO_NAME = aoi$GEO_NAME)
  #add to the big dataframe
  nhood_canopy_lu <- bind_rows(nhood_canopy_lu, cur_canopy)
}

#post-processing
nhood_canopy_lu <- nhood_canopy_lu %>% 
  select(-Water) %>% 
  mutate(across(Agriculture:Woodlands, round)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
#delete any columns with only NA values
nhood_canopy_lu <- nhood_canopy_lu[, colSums(is.na(nhood_canopy_lu)) != nrow(nhood_canopy_lu)]

  
#calculate the % of total canopy on each land use category for ctu

#create ctu group canopy lu dataframe
ctu_canopy_lu <- data.frame(matrix(nrow = 0, ncol=17))
colnames(ctu_canopy_lu) = c("GEO_NAME", unlist(unique(lui$LANDUSE_CA)))
ctu_canopy_lu$GEO_NAME = as.character(ctu_canopy_lu$GEO_NAME)

ctu_geo_proj <- st_transform(ctu_geo, 4269)

for (i in 1:nrow(ctu_geo_proj)) {
  print(paste(i, "/", nrow(ctu_geo_proj)))
  #grab the vector data for the area
  aoi <- ctu_geo_proj[i,]
  #mask the tree and lui rasters to the correct area
  tree_mask_crop <- crop(tree_mask, aoi, mask=TRUE)
  lui_crop <- crop(lui, aoi, mask=TRUE)
  
  #find the amount of canopy on each type of land
  cur_canopy <- zonal(tree_mask_crop, lui_crop, fun="sum", na.rm=TRUE)
  #Subtract any canopy on water (due to sentinel 10m resolution) from the total %s calculated
  if(length(filter(cur_canopy, LANDUSE_CA == 'Water')$trees) != 0) {
    total <- sum(cur_canopy$trees) - filter(cur_canopy, LANDUSE_CA == 'Water')$trees
  } else {
    total <- sum(cur_canopy$trees)
  }
  
  cur_canopy <- cur_canopy %>%
    mutate(perc = trees/total*100) %>%
    select(-trees) %>%
    spread(LANDUSE_CA, perc) %>%
    mutate(GEO_NAME = aoi$GEO_NAME)
  #add to the big dataframe
  ctu_canopy_lu <- bind_rows(ctu_canopy_lu, cur_canopy)
}

#post-processing
ctu_canopy_lu <- ctu_canopy_lu %>% 
  select(-Water) %>%
  mutate(across(Agriculture:Woodlands, round)) %>% 
  mutate(across(Agriculture:Woodlands, ~ case_when(.x < 5 ~ NA, .x > 5 ~ .x)))
  #delete any columns with only NA values
ctu_canopy_lu <- ctu_canopy_lu[, colSums(is.na(ctu_canopy_lu)) != nrow(ctu_canopy_lu)]
  
  
 
save(bg_canopy_lu, ctu_canopy_lu, nhood_canopy_lu, file = paste0(here::here(), "/data-raw/landuse_data.rda"))

plant_opp <- zonal(opp_map, r, fun="sum", na.rm=TRUE) %>% 
  arrange(desc(focal_sum)) %>%
  drop_na() %>% 
  mutate(perc = round((focal_sum*100/(sum(plant_opp$focal_sum)*100))*100)) %>% 
  filter(perc >= 5)


a <- bg_canopy_lu[70,] %>% 
  pivot_longer(Agriculture:Woodlands, names_to = "class") %>% 
  drop_na() %>% 
  arrange(value) %>%
  mutate(class=factor(class, levels=class))   # This trick update the factor levels

  # Plot
ggplot(a, aes(x=class, y=value)) +
  geom_point() + 
  geom_segment( aes(x=class, xend=class, y=0, yend=value)) +
  coord_flip()

#replace with labels
levels(opp_map) <- data.frame(id=1:3, cover=c("Low Canopy", "Moderate Canopy", "High Canopy"))
is.factor(opp_map)

#visualize
plot(aoi$geometry)
plot(opp_map, col=rev(grDevices::hcl.colors(3, "viridis")), add=TRUE)

#calculate some stats
total <- st_area(aoi)
low_perc <- sum(opp_map[opp_map$cover == "Low Canopy"])*100
mod_perc <- sum(opp_map[opp_map$cover == "Moderate Canopy"])*100
high_perc <- sum(opp_map[opp_map$cover == "High Canopy"])*100
print(paste0("total planting opportunities: ", round(sum(low_perc, mod_perc, high_perc)/total*100, 2), "%"))
print(paste0("low: ", round(low_perc/total*100, 2), "%"))
print(paste0("moderate: ", round(mod_perc/total*100, 2), "%"))
print(paste0("high: ", round(high_perc/total*100, 2), "%"))

```

```{r}
df <- ctu_canopy_lu[27,] %>% 
  pivot_longer(Agriculture:Woodlands, names_to = "class") %>% 
  drop_na() %>% 
  arrange(value) %>%
  mutate(class=factor(class, levels=class))   # This trick update the factor levels

        ggplot(df, aes(x=class, y=value)) +
        geom_segment( aes(x=class, xend=class, y=0, yend=value)) +
        geom_point(color=councilR::colors$cdGreen, stroke=3) +
        coord_flip() +
        councilR::theme_council() +
        theme(
            panel.grid.minor = element_blank(),
            panel.grid.major = element_blank(),
            strip.placement = "outside",
            axis.title.y = element_text(
              angle = 90,
              vjust = 3
            ),
            plot.margin = margin(7, 7, 7, 7),
            axis.line = element_line(),
            axis.ticks = element_line(),
            axis.text.y = element_text(vjust = .5, hjust = 1, size=10),
            plot.caption = element_text(
              size = rel(1),
              colour = "grey30",
              vjust = -2
            )
          ) +
        labs(
            x = "Land use category", y = "Percentage canopy",
            caption = # expression(italic(
              "Source: Analysis of Sentinel-2 satellite imagery (2021)\nand Dane County land use inventory (2020)" # ))
          ) 
```





```{r}
lui <- st_read(paste0(here::here(), "/data-raw/LUI_2020_Dane_County/LUI_2020_EXPORT_for_CARPC.gdb"))
#maybe I should just say where current canopy & opportunities are on the primary land use categories instead of doing a more complex catagorization...
st_write(lui, paste0(here::here(), "/data-raw/LUI_2020_Dane_County/LUI_2020_EXPORT_for_CARPC.csv"), layer_options = "GEOMETRY=AS_WKT")
#don't use: water

#not sure: communication/utilites, open land

#recreation (subcategory = public)
public_rec <- filter(lui, grepl('Way', LANDUSE_SUBCATEGORY))
#residential, agriculture farm buildings (subcategory)
residential <- filter(lui, LANDUSE_CATEGORY == 'Residential')
#commercial, industrial, recreation (subcategory = private), all agriculture != subcategory 'agriculture farm buildings' 
private_nonres <- filter(lui, LANDUSE_CATEGORY == 'Commercial' | 'Industrial')
#school land as own variable?

                                                                 
                                             
#"Institutional/Governmental" "Woodlands"                  "Mineral Extraction"        
#[13] "Vacant Subdivided Land"     "Fallow Farm Field"          "Cemetery"                   "Underconstruction"  

```

