---
title: "Growing Shade Report"
date: "`r format(Sys.time(), '%d %B %Y')`"

params:
  param_geo: NA
  param_area: NA
  param_equitypara: NA
  param_treeplot: NA
  param_ranktext: NA
  param_rankplot: NA
  param_prioritytable: NA
  param_equityplot: NA
  param_tempplot: NA
  param_equitytext: NA
  para_heattext: NA
  param_landuseplot: NA
  param_landusetext: NA
  param_opptext: NA
  param_oppplot: NA

urlcolor: blue
always_allow_html: true


---



```{css, echo=FALSE, include = F, eval =F}
caption {
    font-size: .9em;
    font-color: #A9A9A9
}
```



```{r assignments, echo=F, message=F, warning=F, results='hide'}
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

knitr::opts_chunk$set(echo = TRUE, fig.align="center", echo = F)

# fancyname <- params$param_area


```


# Report for `r params$param_area`


```{r map, echo = F, message = F, warning = F, fig.alt = "Outline of the Twin Cities region, the area which this report was made for is mapped in green.", fig.align="right", out.width= "43%", out.extra='style="float:right; padding:10px"'}
#, fig.cap = "\\label{fig:map}The geographic location of the selected area. The selected area is highlighed in green."
if(params$param_geo == "blockgroups") {
  region_outline %>% ggplot() + geom_sf(fill = NA, color = "grey80") +
    geom_sf(data = filter(mn_bgs, fancyname == params$param_area), color = councilR::colors$cdGreen, fill = councilR::colors$cdGreen) +
    theme_void()
  } else if (params$param_geo == "ctus") {
     region_outline %>% ggplot() + geom_sf(fill = NA, color = "grey80") +
    geom_sf(data = filter(ctu_list, GEO_NAME == params$param_area), color = councilR::colors$cdGreen, fill = councilR::colors$cdGreen) +
    theme_void()
  } else if (params$param_geo == "nhood") {
     region_outline %>% ggplot() + geom_sf(fill = NA, color = "grey80") +
    geom_sf(data = filter(nhood_list, GEO_NAME == params$param_area), color = councilR::colors$cdGreen, fill = councilR::colors$cdGreen) +
    theme_void()}

# Growing Shade was developed for planners and foresters, but is available to anyone who wants this information.
# 
# Trees provide critical ecosystem services and are important components of the human, natural and built environments. Enhancing and maintaining tree canopy cover is an actionable step to create healthy places for people to live and a resilient future.

# This report synthesizes and summarizes data about trees, people, and the built environments. Understanding the tree canopy within this larger context is important for prioritization and planning efforts. Please review "what's next for Growing Shade" or contact us at SOME EMAIL HERE if this report does not address your data needs - we welcome the feedback and may be able to accommodate requests.

```


This report was generated from the Growing Shade interactive resource produced by the Tree Canopy Collaborative.

Trees help communities thrive. Community-wide benefits from trees include cooler temperatures, improved air quality, cleaner stormwater, improved mental health, and enhanced biodiversity. Enhancing tree canopy cover is an actionable step to create healthy and resilient places. The Growing Shade Project provides data and resources to inform tree planting and canopy preservation for Dane County.

Trees connect to regional issues and priorities. Public health, environmental justice, climate change, and natural resource conservation all intersect with the tree canopy. Tree planting efforts and ongoing maintenance will benefit from new perspectives, rigorous science, and thoughtful synthesis. Growing Shade was developed for planners and foresters but is available to anyone who wants this information. Stakeholder engagement remains essential to understand community-specific concerns and opportunities for enhancing, preserving, and maintaining the tree canopy.




## Prioritization

`r params$param_ranktext`

Values for all potential variables can be obtained by downloading the raw data from the "mapping tool" tab.

```{r priorityplot, fig.align="center", out.width= "80%", fig.asp = .5, fig.alt = "Figure showing the priority ranking (climate change, conservation, environmental justice, public health) for all block groups within the selected geography."}
params$param_rankplot
```


<br>

```{r priority-table, results='markup', tidy = F}
knitr::kable(params$param_prioritytable,
             booktabs = TRUE) %>%
             # caption = "Average values for the selected area as compared to the region-wide average. Variables shown are those used in prioritization layer. Values for all potential variables can be obtained by downloading the raw data from the Growing Shade Mapping Tool.") %>%
  # kableExtra::kable_styling(full_width = F) %>%
    kableExtra::kable_classic("striped", full_width = F, html_font = "Cambria")# %>%
  # kableExtra::footnote(general = "Average values for the selected area as compared to the region-wide average. Variables shown are those used in prioritization layer. Values for all potential variables can be obtained by downloading the raw data from the Growing Shade Mapping Tool.",
           # general_title = "")
```


## Tree canopy

`r params$param_equitypara`


```{r treeplot, fig.align="center", out.width= "70%", fig.asp = .4, fig.alt = "Figure showing the distribution of tree canopy across the region and within the selected geography."}
params$param_treeplot
# Census block groups have different areas and may overlap with other geographies, thus the existing tree canopy cover in the selected area may not be the mean of the block group canopy covers.

```

```{r include = F, eval = F}
### Methodology note

# The method used to detect tree canopy varies across tools. The 30% goal of canopy coverage and values about current canopy coverage in Growing Shade may not align with goals or values presented in other tools. In Growing Shade, we prioritize using current data. Current data is particularly important as our region continues to feel the impact of emerald ash borer on the tree canopy. 
# 
# To get current (near real-time) data, Growing Shade leverages metrics of greenness (normalized difference vegetation index, NDVI) and seasonal patterns in plant growth (phenology) to detect trees from Sentinel-2 satellite data. Sentinel collects data approximately every 10 days, and our method does not need to rely on older data (for instance, LiDAR data for our region is from 2011 and out-of-date). While Sentinel's temporal resolution is excellent, the satellite data has a spatial resolution of 10 meters squared. This is often a bigger area than tree canopies. 
# 
# When comparing tree canopy coverage from Sentinel with a more spatially accurate (but less temporally accurate) 1 meter squared landcover data set, we found that Sentinel tree cover has high correlation with 1 meter landcover, but Sentinel detects about twice as much tree canopy. Essentially, this means that we are detecting areas with at least 50% tree canopy coverage. We re-scaled our data using this relationship to improve the clarity of messaging. Thus, another way to view the goal of 30% canopy cover is to have 60% of land area covered by at least 50% tree canopy. 
```


## Race & income disparities

`r params$param_equitytext`

```{r include = T, eval = F}
# Research shows that trees are unevenly distributed across communities. In particular, areas with a high percent of the population identifying as a person of color or low-income populations have less tree canopy (<a href = "https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0249715" target="_blank">MacDonald 2021</a>) as do historically redlined areas (<a href = "https://www.npr.org/2020/01/14/795961381/racist-housing-practices-from-the-1930s-linked-to-hotter-neighborhoods-today" target = "_blank">NPR 2020</a>, <a href = "https://www.nature.com/articles/s42949-021-00022-0" target = "_blank">Locke et al. 2021</a>, <a href = "https://www.sciencedirect.com/science/article/abs/pii/S0277953619307531?via%3Dihub" target = "_blank">Namin et al. 2020</a>). Addressing inequity in tree canopy cover may reduce heat-related deaths by up to 25% (<a href = "https://www.fs.fed.us/nrs/pubs/jrnl/2021/nrs_2021_paramita_001.pdf" target = "_blank">Sinha 2021</a>). 

# (358 / 2)/(.12*620961) * 100
# Results indicate that approximately 251 deaths (95% confidence interval of 143 to 358) can be avoided in the city of Baltimore (total population of 620,961) with a 10% increase in tree cover. We also find that more than half of the reduced mortality impacts are for the population over the age of 65 (which is about 12% of the total population)
```

```{r equityplot, fig.align="center", out.width= "70%", fig.asp = .9, fig.alt = "Figure showing the trends between tree canopy and median household income and the percent of population identifying as a person of color."}
params$param_equityplot
```


```{r old_threats, include = F, eval = F}
### Methodology note

# The normalized difference vegetation index (NDVI) ranges from -1 to 1. Areas with bare ground, or brown vegetation have NDVI values closer to zero. Areas with lots of greenness have values closer to one. The calculation for land surface temperature *does* rely on information about NDVI, so these two variables are not fully independent from each other. Land surface temperature is calculated using other data as well. The NDVI data shown here comes from the Sentinel-2 satellite (resolution of 10 x 10 meters) and is from year 2021. Land surface temperature shown here comes from data collected by the Landsat-8 satellite (resolution of 30 x 30 meters) and is from year 2016.

## Threats

# `r params$param_otherparea`

```

<br>

For information on canopy resources or details about the methodology of this report, please visit: https://capital-area-rpc.shinyapps.io/growing_shade-main/
```{r other-resources-methods}
#uncomment below to include full text from these pages
#includeMarkdown(system.file("app/www/otherresources2.md", package = "planting.shade"))
#includeMarkdown(system.file("app/www/methods2.md", package = "planting.shade"))

```


<br><br>






